Modern Byte Pair Tokenizers are Zipfian

### 2025-05-14 writeup/paper.tex
whole draft lol

@@

\title{Are Modern Byte Pair Tokenizers Zipfian?}

@@

\begin{document}
\maketitle
\begin{abstract}
This document is a supplement to the general instructions for *ACL authors. It contains instructions for using the \LaTeX{} style files for ACL conferences.
The document itself conforms to its own specifications, and is therefore an example of what your manuscript should look like.
These instructions should be used both for papers submitted for review and for final versions of accepted papers.
\end{abstract}

@@

These instructions are for authors submitting papers to *ACL conferences using \LaTeX. They are not self-contained. All authors must follow the general instructions for *ACL proceedings,\footnote{\url{http://acl-org.github.io/ACLPUB/formatting.html}} and this document contains additional instructions for the \LaTeX{} style files.

The templates include the \LaTeX{} source of this document (\texttt{acl\_latex.tex}),
the \LaTeX{} style file used to format it (\texttt{acl.sty}),
an ACL bibliography style (\texttt{acl\_natbib.bst}),
an example bibliography (\texttt{custom.bib}),
and the bibliography for the ACL Anthology (\texttt{anthology.bib}).

\section{Engines}

To produce a PDF file, pdf\LaTeX{} is strongly recommended (over original \LaTeX{} plus dvips+ps2pdf or dvipdf).
The style file \texttt{acl.sty} can also be used with
lua\LaTeX{} and
Xe\LaTeX{}, which are especially suitable for text in non-Latin scripts.
The file \texttt{acl\_lualatex.tex} in this repository provides
an example of how to use \texttt{acl.sty} with either
lua\LaTeX{} or
Xe\LaTeX{}.

\section{Preamble}

The first line of the file must be
\begin{quote}
\begin{verbatim}
\documentclass[11pt]{article}
\end{verbatim}
\end{quote}

To load the style file in the review version:
\begin{quote}
\begin{verbatim}
\usepackage[review]{acl}
\end{verbatim}
\end{quote}
For the final version, omit the \verb|review| option:
\begin{quote}
\begin{verbatim}
\usepackage{acl}
\end{verbatim}
\end{quote}

To use Times Roman, put the following in the preamble:
\begin{quote}
\begin{verbatim}
\usepackage{times}
\end{verbatim}
\end{quote}
(Alternatives like txfonts or newtx are also acceptable.)

Please see the \LaTeX{} source of this document for comments on other packages that may be useful.

Set the title and author using \verb|\title| and \verb|\author|. Within the author list, format multiple authors using \verb|\and| and \verb|\And| and \verb|\AND|; please see the \LaTeX{} source for examples.

By default, the box containing the title and author names is set to the minimum of 5 cm. If you need more space, include the following in the preamble:
\begin{quote}
\begin{verbatim}
\setlength\titlebox{<dim>}
\end{verbatim}
\end{quote}
where \verb|<dim>| is replaced with a length. Do not set this length smaller than 5 cm.

\section{Document Body}

\subsection{Footnotes}

Footnotes are inserted with the \verb|\footnote| command.\footnote{This is a footnote.}

\subsection{Tables and figures}

See Table~\ref{tab:accents} for an example of a table and its caption.
\textbf{Do not override the default caption sizes.}

\begin{table}
  \centering
  \begin{tabular}{lc}
    \hline
    \textbf{Command} & \textbf{Output} \\
    \hline
    \verb|{\"a}|     & {\"a}           \\
    \verb|{\^e}|     & {\^e}           \\
    \verb|{\`i}|     & {\`i}           \\
    \verb|{\.I}|     & {\.I}           \\
    \verb|{\o}|      & {\o}            \\
    \verb|{\'u}|     & {\'u}           \\
    \verb|{\aa}|     & {\aa}           \\\hline
  \end{tabular}
  \begin{tabular}{lc}
    \hline
    \textbf{Command} & \textbf{Output} \\
    \hline
    \verb|{\c c}|    & {\c c}          \\
    \verb|{\u g}|    & {\u g}          \\
    \verb|{\l}|      & {\l}            \\
    \verb|{\~n}|     & {\~n}           \\
    \verb|{\H o}|    & {\H o}          \\
    \verb|{\v r}|    & {\v r}          \\
    \verb|{\ss}|     & {\ss}           \\
    \hline
  \end{tabular}
  \caption{Example commands for accented characters, to be used in, \emph{e.g.}, Bib\TeX{} entries.}
  \label{tab:accents}
\end{table}

As much as possible, fonts in figures should conform
to the document fonts. See Figure~\ref{fig:experiments} for an example of a figure and its caption.

Using the \verb|graphicx| package graphics files can be included within figure
environment at an appropriate point within the text.
The \verb|graphicx| package supports various optional arguments to control the
appearance of the figure.
You must include it explicitly in the \LaTeX{} preamble (after the
\verb|\documentclass| declaration and before \verb|\begin{document}|) using
\verb|\usepackage{graphicx}|.

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{example-image-golden}
  \caption{A figure with a caption that runs for more than one line.
    Example image is usually available through the \texttt{mwe} package
    without even mentioning it in the preamble.}
  \label{fig:experiments}
\end{figure}

\begin{figure*}[t]
  \includegraphics[width=0.48\linewidth]{example-image-a} \hfill
  \includegraphics[width=0.48\linewidth]{example-image-b}
  \caption {A minimal working example to demonstrate how to place
    two images side-by-side.}
\end{figure*}

\subsection{Hyperlinks}

Users of older versions of \LaTeX{} may encounter the following error during compilation:
\begin{quote}
\verb|\pdfendlink| ended up in different nesting level than \verb|\pdfstartlink|.
\end{quote}
This happens when pdf\LaTeX{} is used and a citation splits across a page boundary. The best way to fix this is to upgrade \LaTeX{} to 2018-12-01 or later.

\subsection{Citations}

\begin{table*}
  \centering
  \begin{tabular}{lll}
    \hline
    \textbf{Output}           & \textbf{natbib command} & \textbf{ACL only command} \\
    \hline
    \citep{Gusfield:97}       & \verb|\citep|           &                           \\
    \citealp{Gusfield:97}     & \verb|\citealp|         &                           \\
    \citet{Gusfield:97}       & \verb|\citet|           &                           \\
    \citeyearpar{Gusfield:97} & \verb|\citeyearpar|     &                           \\
    \citeposs{Gusfield:97}    &                         & \verb|\citeposs|          \\
    \hline
  \end{tabular}
  \caption{\label{citation-guide}
    Citation commands supported by the style file.
    The style is based on the natbib package and supports all natbib citation commands.
    It also supports commands defined in previous ACL style files for compatibility.
  }
\end{table*}

Table~\ref{citation-guide} shows the syntax supported by the style files.
We encourage you to use the natbib styles.
You can use the command \verb|\citet| (cite in text) to get ``author (year)'' citations, like this citation to a paper by \citet{Gusfield:97}.
You can use the command \verb|\citep| (cite in parentheses) to get ``(author, year)'' citations \citep{Gusfield:97}.
You can use the command \verb|\citealp| (alternative cite without parentheses) to get ``author, year'' citations, which is useful for using citations within parentheses (e.g. \citealp{Gusfield:97}).

A possessive citation can be made with the command \verb|\citeposs|.
This is not a standard natbib command, so it is generally not compatible
with other style files.

\subsection{References}

\nocite{Ando2005,andrew2007scalable,rasooli-tetrault-2015}

The \LaTeX{} and Bib\TeX{} style files provided roughly follow the American Psychological Association format.
If your own bib file is named \texttt{custom.bib}, then placing the following before any appendices in your \LaTeX{} file will generate the references section for you:
\begin{quote}
\begin{verbatim}
\bibliography{custom}
\end{verbatim}
\end{quote}

You can obtain the complete ACL Anthology as a Bib\TeX{} file from \url{https://aclweb.org/anthology/anthology.bib.gz}.
To include both the Anthology and your own .bib file, use the following instead of the above.
\begin{quote}
\begin{verbatim}
\bibliography{anthology,custom}
\end{verbatim}
\end{quote}

Please see Section~\ref{sec:bibtex} for information on preparing Bib\TeX{} files.

\subsection{Equations}

An example equation is shown below:
\begin{equation}
  \label{eq:example}
  A = \pi r^2
\end{equation}

Labels for equation numbers, sections, subsections, figures and tables
are all defined with the \verb|\label{label}| command and cross references
to them are made with the \verb|\ref{label}| command.

This an example cross-reference to Equation~\ref{eq:example}.

\subsection{Appendices}

Use \verb|\appendix| before any appendix section to switch the section numbering over to letters. See Appendix~\ref{sec:appendix} for an example.

\section{Bib\TeX{} Files}
\label{sec:bibtex}

Unicode cannot be used in Bib\TeX{} entries, and some ways of typing special characters can disrupt Bib\TeX's alphabetization. The recommended way of typing special characters is shown in Table~\ref{tab:accents}.

Please ensure that Bib\TeX{} records contain DOIs or URLs when possible, and for all the ACL materials that you reference.
Use the \verb|doi| field for DOIs and the \verb|url| field for URLs.
If a Bib\TeX{} entry has a URL or DOI field, the paper title in the references section will appear as a hyperlink to the paper, using the hyperref \LaTeX{} package.

\section*{Limitations}

Since December 2023, a "Limitations" section has been required for all papers submitted to ACL Rolling Review (ARR). This section should be placed at the end of the paper, before the references. The "Limitations" section (along with, optionally, a section for ethical considerations) may be up to one page and will not count toward the final page limit. Note that these files may be used by venues that do not rely on ARR so it is recommended to verify the requirement of a "Limitations" section and other criteria with the venue in question.

\section*{Acknowledgments}

This document has been adapted
by Steven Bethard, Ryan Cotterell and Rui Yan
from the instructions for earlier ACL and NAACL proceedings, including those for
ACL 2019 by Douwe Kiela and Ivan Vuli\'{c},
NAACL 2019 by Stephanie Lukin and Alla Roskovskaya,
ACL 2018 by Shay Cohen, Kevin Gimpel, and Wei Lu,
NAACL 2018 by Margaret Mitchell and Stephanie Lukin,
Bib\TeX{} suggestions for (NA)ACL 2017/2018 from Jason Eisner,
ACL 2017 by Dan Gildea and Min-Yen Kan,
NAACL 2017 by Margaret Mitchell,
ACL 2012 by Maggie Li and Michael White,
ACL 2010 by Jing-Shin Chang and Philipp Koehn,
ACL 2008 by Johanna D. Moore, Simone Teufel, James Allan, and Sadaoki Furui,
ACL 2005 by Hwee Tou Ng and Kemal Oflazer,
ACL 2002 by Eugene Charniak and Dekang Lin,
and earlier ACL and EACL formats written by several people, including
John Chen, Henry S. Thompson and Donald Walker.
Additional elements were taken from the formatting instructions of the \emph{International Joint Conference on Artificial Intelligence} and the \emph{Conference on Computer Vision and Pattern Recognition}.

@@

\appendix

\section{Example Appendix}
\label{sec:appendix}

This is an appendix.

>>>>

% Jack!
% \usepackage{amsmath}

@@

\title{Modern Byte Pair Tokenizers are Zipfian}

@@

\begin{document}
\maketitle
\begin{abstract}
A majority of large language models ingest word fragments called \textit{tokens} produced by a data compression algorithm known as byte pair encoding. This algorithm groups high-frequency letter pairings in natural language into individual units. A natural question is whether the tokens produced by this pairing process deviate significantly from the source language's frequency distribution. Zipf famously showed that many natural language's frequency distribution follows a power law, commonly known as Zipf's Law. We examine two modern tokenizer's adherence to Zipf's law at the token level. We demonstrate that these tokenizers are Zipfian on two corpuses, and speculate as to why this is.
\end{abstract}

@@

George Zipf in his hallmark \textit{The Psycho-Biology of Language} \cite{Zip35} introduced the remarkable trend that the word frequencies in many human languages exhibit the same power law distribution. Specifically, Zipf said that the frequency  that a word appears in some large corpus is proportional to the word's frequency rank. Mathematically, Zipf's laws states

\begin{equation}
    \mbox{word frequency} \propto \frac{1}{\mbox{word rank}}.
\end{equation}

A more descriptive model of word frequency that is more commonly referenced in linguistics literature is the Zipf-Mandelbrot distribution

\begin{equation}
    \mbox{word frequency} \propto \frac{1}{(\mbox{word rank} + b)^a}.
    \label{eq:zipfmandelbrot}
\end{equation}

where $a,b$ are fitted parameters. We say a distribution that follows the trend in Equation \ref{eq:zipfmandelbrot} with $a \approx 1$ is \textit{Zipf distributed}, or simply \textit{Zipfian}.

The accuracy of the trend in Equation \ref{eq:zipfmandelbrot} has been examined in $10^8$ English words in \cite{cancho2000}, over 50 languages in \cite{yu2018zipfslaw50languages}, and written-versus-spoken corpuses in \cite{lin2015scalinglawshumanspeech}. Each of these studies demonstrate that Zipf's law is generally exhibitted for common and somewhat-uncommon words, but rare words (words with high rank) appear less frequently than predicted. This deviation is shown to be statistically significant, and appears as two trendlines in the log-log plot. In English corpuses, this transition is found around the $10^5$\textit{-th} ranked word.\footnote{Some of these studies also find that extremely common words appear slighly more commonly than the Zipfian prediction.} The authors further explore the linguistic relavence and universality of these multiple regimes of words.

Nearly a century after Zipf's discovery, large language models (LLMs) generate text comparable to human communication. However, unlike humans, LLMs digest text using a fixed vocabulary of word fragments called \textit{tokens}. The mapping between natural language and tokens is most commonly computed using the \textit{byte pair encoding algorithm} \cite{bpegage} over some large training corpus. Byte pair encoding identifies the most frequent pair of letters, and creates a new token in the "token vocabulary" to replace that pair. Iterating this procedure until some fixed vocabulary size is reached creates the vocabulary of tokens. Finally, any sequence that does not map to one of the derived tokens is labelled with the \texttt{<unk>} token, standing for unknown.

When considering these ideas in tandem, a natural question arises: given a corpus that appears Zipfian, how does the tokenization process affect this trend?

\section{Methods}

To explore this question, we choose two modern tokenizers to analyze. We choose the tokenizer for the RoBERTa language model \cite{liu2019robertarobustlyoptimizedbert}, as the training data is publicly known. This allows us to conduct frequency analyses on corpuses that are on and off-distribution for the tokenizer. We also choose to compare our results with the tokenizer for the GPT-4 language model \cite{openai2024gpt4technicalreport} as a representative of an industry-grade tokenizer.

We choose two corpuses based on RoBERTa's training data. We choose the 4.4GB bookscorpus dataset \cite{zhu2015aligningbooksmoviesstorylike}, as RoBERTa's tokenizer was trained (in-part) on this corpus. For our off-distribution corpus, we choose the 5.6GBtying MiniPile dataset \cite{kaddour2023minipilechallengedataefficientlanguage}, which was released after the RoBERTa model was released. Note that it is likely that the GPT-4 tokenizer was trained (in-part) on both of these corpuses.

We compute the word and token frequency for both tokenizers on each corpus, where we define a word as anything separated by a space. We then compute the data's Kolmogorov-Smirnov goodness-of-fit statistic for the Zipf distribution, and compare to that for an exponential and log-normal distribution. We do not remove any tokens from consideration unless otherwise stated.

\section{Results}

We find that token vocabularies are remarkably Zipfian, excluding extremely rare tokens. This holds true for both tokenizers and both corpuses studied, including out-of-distribution text. In the log-log plots in Figure \ref{fig:experiments}, we show sharp declines from the Zipfian trend only for the final few tokens in the RoBERTa tokenizer for both corpuses and the GPT-4 tokenizer for the MiniPile corpus. However, the GPT-4 tokenizer trend on the bookscorpus seems to deviate more dramatically, similar to the multiple regime studies.

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{../visualizations/bookscorpusfreq.png}
  \includegraphics[width=\columnwidth]{../visualizations/minipilecorpusfreq.png}
  \caption{Log-log plots of word \& token rank vs word \& token frequency}
  \label{fig:experiments}
\end{figure}

The lowest rank words and tokens for each tokenizer and each corpus, including control tokens and punctuation, are summarized in Table \ref{tbl:token ranks}.

\begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|c|c|c|}
        \hline
        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
        \hline
        1 & . & the & <s> & ␣ & `` & , \\
        \hline
        2 & , & of & \textbackslash n & \textbackslash n & i & ␣the \\
        \hline
        3 & the & and & <\textbackslash s> & . & he & . \\
        \hline
    \end{tabular}
    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn MiniPile, including control tokens and punctuation}
    \label{tbl:token ranks}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|c|c|c|}
        \hline
        \textbf{Rank} & \multicolumn{2}{c|}{\textbf{Words}} & \multicolumn{2}{c|}{\textbf{RoBERTa}} & \multicolumn{2}{c|}{\textbf{GPT-4}}  \\
        \hline
        1 & the & the & the & the & i & the \\
        \hline
        2 & to & of & to & of & he & of \\
        \hline
        3 & i & and & and & and & she & and \\
        \hline
    \end{tabular}
    \caption{Summary of most common words \& tokens for each corpus, where the left subcolumn is bookscorpus and the right subcolumn MiniPile, \textit{not} including control tokens and punctuation}
    \label{tbl:token ranks no control}
\end{table}

The p-values for the Kolmogorov-Smirnov Tests for each considered distribution are summarized in Table \ref{tbl:pvals}.

\begin{table}
    \centering
    \begin{tabular}{|c|c|c|}
        \hline
        \textbf{Distribution} & \textbf{bookscorpus} & \textbf{MiniPile} \\
        \hline
        Zipf &  &  \\
        \hline
        Exponential &  &  \\
        \hline
        log-normal &  &  \\
        \hline
    \end{tabular}
    \caption{RoBERTa (GPT-4) tokenizer's Kolmogorov-Smirnov test p-values}
    \label{tbl:pvals}
\end{table}

\section{Conclusions}

Despite

\section{Future Work}

TODO

How do token vocabularies differ from english vocabularies? Fixed vocabulary size, catchall token, functional tokens

@@

% \appendix

% \section{Example Appendix}
% \label{sec:appendix}

% This is an appendix.
